Nested Stochastic Block Models Applied to the Analysis of Single Cell Data

### 2021-07-15 manuscript.tex
Removed TIAN0

\subsection*{Absence of information is correctly handled by nSBM}

One of the most relevant difference between SBM and other methods to cluster single cells is that it relies on robust statistical modelling. In this sense, the number of groups identified strictly mirrors the amount of information contained in the data. An important consequence is that absence of information can be properly handled. To show this property we performed a simple experiment on a randomised \emph{k}NN graph. We collected data for 3k PBMC (available as preprocessed data in \texttt{scanpy}, Fig. \ref{FigureRandom}A) and shuffled the edges of the prebuilt \emph{k}NN graph, this to keep the general graph properties unchanged. We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0733$, $p=0.703$). We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups, whereas nSBM does not identify any cell group (Fig. \ref{FigureRandom}B).

@@

This experiment is a deliberate extreme case. The quality of grouping proposed by a standard approach can be disputed in many ways, and the UMAP embedding indeed reflects the absence of any information. Nevertheless, real-world data may include an unknown amount of random noise. Hence, It is important to identify cell groups that are not artefacts arising from processing and that do reflect the information contained in the dataset.

@@

To benchmark \texttt{schist}, we tested it on scRNA-seq mixology data \cite{Tian_2019}, a dataset explicitly developed to benchmark single cell analysis tools without the need to simulate data. In particular, we used the mixture of 5 cell lines profiled with Chromium 10x platform. At a first evaluation of the UMAP embedding, all lines appear well separated. Only the lung cancer line H1975 shows a considerable degree of heterogeneity and appears to be split into two cell groups (Fig. \ref{FigureTian}A). Using default parameters, \texttt{schist} is able to identify correct cell groups ($ARI=0.829$), with a further split in H2228 cell line (Fig. \ref{FigureTian}B), whereas Leiden method clusters the dataset into 10 groups ($ARI=0.549$, Fig. \ref{FigureTian}C). The nSBM solution correctly identifies H1975 groups as a single entity at level 1 of the hierarchy. We then sought to check if an independent agglomerative method, SCCAF \cite{miao_2020}, was able to recover cell line groupings starting from both partition schemes. Given the ground truth, the cell lines, SCCAF is able to assess the maximal accuracy that can be achieved in the dataset (0.992). When trained with this target accuracy, SCCAF precisely reconstructs the original cell line annotations starting from \texttt{schist} partitions with high accuracy (Fig. \ref{FigureTian}D). When Leiden partitions are set as input, SCCAF merges H2228 and HCC827 cells into a single cluster and keeps H1975 cells split into two groups (Fig. \ref{FigureTian}E), highlighting potential limitations of this approach. In order to perform a fair and complete experiment, we also tested SCCAF with the partitions scheme at the deepest level of the nSBM hierarchy (38 groups, Fig. \ref{Tian_NSBM0}A), for which we observed the inability to perform a significant cell grouping (Fig. \ref{Tian_NSBM0}B).

>>>>

\subsection*{\texttt{schist} is robust to data perturbation}

One of the most relevant difference between the SBM and other methods to cluster single cells is that it relies on robust statistical modelling. In this sense, the number of groups identified strictly mirrors the amount of information contained in the data. An important consequence is that absence of information \textcolor{red}{(\emph{i.e.} maximal entropy)} can be properly handled. To show this property we performed a simple experiment on a randomised \emph{k}NN graph. We collected data for 3k PBMC (available as preprocessed data in \texttt{scanpy}, Fig. \ref{FigureRandom}A) and shuffled the edges of the prebuilt \emph{k}NN graph, this to keep the general graph properties unchanged. We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0733$, $p=0.703$). We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups, whereas nSBM does not identify any cell group \textcolor{red}{at default resolution} (Fig. \ref{FigureRandom}B).

@@

\textcolor{red}{Only by reducing resolution to $\gamma \leq 0.5$ we were able to obtain a single partition by modularity (Fig. \ref{Modularity_Random_Res}). Of course,} this experiment is a deliberate extreme case. The quality of grouping proposed by a standard approach can be disputed in many ways, and the UMAP embedding indeed reflects the absence of any information. Nevertheless, real-world data may include an unknown amount of random noise. Hence, It is important to identify cell groups that are not artefacts arising from processing and that do reflect the information contained in the dataset. \textcolor{red}{To understand the impact of random noise on structured data, we considered the same PBMC dataset and added white noise to the normalised counts at increasing levels of $\sigma$, ensuring that the noise level is modeled after the feature-wise distribution of detected genes. We then compared partitions to the original annotation by Adjusted Rand Index ($ARI$), we found that \texttt{schist} is more robust to perturbations and that, again, optimising the modularity results in overestimation of the number of communities at high noise levels (Table \ref{Table_Noise}. Of note, the concordance with original annotations drops at $\sigma \geq 1.5$ for both the approaches.}

@@

To benchmark \texttt{schist}, we tested it on scRNA-seq mixology data \cite{Tian_2019}, a dataset explicitly developed to benchmark single cell analysis tools without the need to simulate data. In particular, we used the mixture of 5 cell lines profiled with Chromium 10x platform. At a first evaluation of the UMAP embedding, all lines appear well separated. Only the lung cancer line H1975 shows a considerable degree of heterogeneity and appears to be split into two cell groups (Fig. \ref{FigureTian}A). Using default parameters, \texttt{schist} is able to identify correct cell groups ($ARI=0.829$), with a further split in H2228 cell line (Fig. \ref{FigureTian}B), whereas Leiden method clusters the dataset into 10 groups ($ARI=0.549$, Fig. \ref{FigureTian}C). The nSBM solution correctly identifies H1975 groups as a single entity at level 1 of the hierarchy. We then sought to check if an independent agglomerative method, SCCAF \cite{miao_2020}, was able to recover cell line groupings starting from both partition schemes. Given the ground truth, the cell lines, SCCAF is able to assess the maximal accuracy that can be achieved in the dataset (0.992). When trained with this target accuracy, SCCAF precisely reconstructs the original cell line annotations starting from \texttt{schist} partitions with high accuracy (Fig. \ref{FigureTian}D). When Leiden partitions are set as input, SCCAF merges H2228 and HCC827 cells into a single cluster and keeps H1975 cells split into two groups (Fig. \ref{FigureTian}E), highlighting potential limitations of this approach. %In order to perform a fair and complete experiment, we also tested SCCAF with the partitions scheme at the deepest level of the nSBM hierarchy (38 groups, Fig. \ref{Tian_NSBM0}A), for which we observed the inability to perform a significant cell grouping (Fig. \ref{Tian_NSBM0}B).

### 2021-07-15 supplementary.tex
Initial work on random data
Removed TIAN0

% Apply footnote and endnote numbering schemes (decimal, lowercase roman etc.)
\renewcommand{\thefootnote}{\arabic{footnote}}
\renewcommand{\theendnote}{\arabic{endnote}}
% Adjust endnotes to use normal-sized numbering and text
\renewcommand{\enotesize}{\normalsize}
\renewcommand\enoteformat{%
  \raggedright
  \leftskip=1.8em
  \makebox[0pt][r]{\theenmark. \rule{0pt}{\dimexpr\ht\strutbox+\baselineskip}}%
}
\renewcommand{\thefigure}{S\arabic{figure}}

\begin{document}
\title{Nested Stochastic Block Models Applied to the Analysis of Single Cell Data}
\author[1,2]{Leonardo Morelli}
\author[1,3]{Valentina Giansanti}
\author[1]{Davide Cittaro}
\affil[1]{Center for Omics Sciences, IRCCS San Raffaele Institute, Milan, Italy}
\affil[2]{Universit\`{a} Vita-Salute San Raffaele, Milan, Italy} 
\affil[3]{Department of Informatics, Systems and Communication, University of Milano-Bicocca, Milan, Italy}
\maketitle
\section*{Supplementary figures}

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Tian_NSBM0.png}
\caption[]{Analysis of a mixture of 5 known cell lines. (A) UMAP  embedding of cell lines from the sc-mixology experiment coloured by the level 0 of the hierarchy proposed by the nested Stochastic Block Model (B) UMAP embedding coloured by the classification made by SCCAF when partitions in (A) are used.}\label{Tian_NSBM0}
\end{figure}
\clearpage
%\begin{figure}[htbp]
%\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S2.png}
%\caption[]{Adjusted Rand Index for different \emph{k}NN graphs after MCMC run. Maximal ARI over all hierarchy level is shown. Darker color indicates higher concordance with the ground truth}\label{MPFigure:9060BBC5-898E-4AA7-A57D-D221167F7250}
%\end{figure}

>>>>

% Apply footnote and endnote numbering schemes (decimal, lowercase roman etc.)
\renewcommand{\thefootnote}{\arabic{footnote}}
\renewcommand{\theendnote}{\arabic{endnote}}
% Adjust endnotes to use normal-sized numbering and text
\renewcommand{\enotesize}{\normalsize}
\renewcommand\enoteformat{%
  \raggedright
  \leftskip=1.8em
  \makebox[0pt][r]{\theenmark. \rule{0pt}{\dimexpr\ht\strutbox+\baselineskip}}%
}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}

\begin{document}
\title{Nested Stochastic Block Models Applied to the Analysis of Single Cell Data}
\author[1,2]{Leonardo Morelli}
\author[1,3]{Valentina Giansanti}
\author[1]{Davide Cittaro}
\affil[1]{Center for Omics Sciences, IRCCS San Raffaele Institute, Milan, Italy}
\affil[2]{Universit\`{a} Vita-Salute San Raffaele, Milan, Italy} 
\affil[3]{Department of Informatics, Systems and Communication, University of Milano-Bicocca, Milan, Italy}
\maketitle

\section*{Supplementary tables}
\begin{table}[h!]
\centering
 \begin{tabular}{||c|c c|c c||}
%  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}
 \hline
\multicolumn{1}{||c|}{\textbf{Noise level ($\sigma$)}} & \multicolumn{2}{c|}{\textbf{Leiden}} & \multicolumn{2}{c||}{\textbf{nSBM level 1}} \\

 & $ARI$ & $N$ & $ARI$ & $N$  \\ 
 \hline\hline
-- & 0.9228 & 8 & 0.8944 & 6 \\
\hline 
0.5 & 0.6249 & 10 & 0.7509  & 10 \\ 
 \hline
 1.0  & 0.5966 & 7 & 0.7269  & 6 \\
 \hline
 1.5  & 0.2833 & 9 &0.3029 & 3 \\
 \hline
 2.0  & 0.1405 & 12 & 0.1961 & 2 \\
 \hline
\end{tabular}
\caption{\textcolor{red}{Effect of random noise }}
\label{Table_Noise}
\end{table}

\clearpage

\section*{Supplementary figures}

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.375\textheight]{Random_Leiden_by_gamma.png}
\caption[]{\textcolor{red}{foobarbaz}}
\label{Modularity_Random_Res}
\end{figure}
\clearpage

%\begin{figure}[htbp]
%\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Tian_NSBM0.png}
%\caption[]{Analysis of a mixture of 5 known cell lines. (A) UMAP  embedding of cell lines from the sc-mixology experiment coloured by the level 0 of the hierarchy proposed by the nested Stochastic Block Model (B) UMAP embedding coloured by the classification made by SCCAF when partitions in (A) are used.}\label{Tian_NSBM0}
%\end{figure}
%\clearpage
%\begin{figure}[htbp]
%\centering
%\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_S2.png}
%\caption[]{Adjusted Rand Index for different \emph{k}NN graphs after MCMC run. Maximal ARI over all hierarchy level is shown. Darker color indicates higher concordance with the ground truth}\label{MPFigure:9060BBC5-898E-4AA7-A57D-D221167F7250}
%\end{figure}
